---
title: "blast_analysis"
author: "m quigg"
date: "2025-12-01"
output: html_document
---
# Who's infected???
Soooooo, I decided to split into "high" and "low" confidence infections because blast is weird.

high= one or more Blast hits matching the genus of the disease
low= one or more hits matching the family of the of the disease

## Import all of the files
```{r}
####import data####
setwd("C:/Users/mquig/Desktop/DED/Data_analysis/disease")
disease=read.csv("final_all_counts_by_individual.csv", header = T)
coords=read.csv("master_list_NA_updated.csv", header = T)
year=read.csv("samples_and_year.csv", header = T)
```

## Take a trip to the library
This script is easy, all we need is tidyverse.
```{r, include=FALSE}
####library####
#install.packages("tidyverse")
library(tidyverse)
```

## Quick organization
```{r}
###data organization
coords=coords %>% 
  select(c(Sample.ID, Lat.NA_corrected., Long.NA_corrected.)) %>% 
  rename("Sample"=Sample.ID) %>% 
  rename("Lat"=Lat.NA_corrected.) %>% 
  rename("Long"=Long.NA_corrected.)

year=year %>% 
  rename("Sample"=ind) %>% 
  rename("Year"=year)
```

##Remove TS09's DED detection
So the phylogney revealed that there were no reads matching to the DED genes from TS09, so I'm considering this a false-detection. It has one Ophiostoma and one Ophiostomatacae so I want to edit those directly.
```{r}
disease[508, 3] = 0
disease[508, 4] = 0
```

## Time to make a dataframe for GIS!
I want a dataframe with a column with "high", "low", or "uninfected" for each disease per individual. I also want a column for coinfections, so I can map that too. It also needs to have the year each tree was sampled, and the coordinates.

I exported the csv file and mapped it in ArcGIS.
```{r, include=FALSE}
###infection status
classified_df <- disease %>%
  mutate(
    DED = case_when(
      Ophiostomataceae >= 1 & Ophiostoma >= 1 & ulmi >= 1 ~ "high",
      Ophiostomataceae >= 1 & ulmi == 0 ~ "low",
      TRUE ~ "uninfected"
    ),
    Phyto = case_when(
      Acholeplasmataceae >= 1 & Phytoplasma >= 1 ~ "high",
      Acholeplasmataceae >= 1 & Phytoplasma == 0 ~ "low",
      TRUE ~ "uninfected"
    ),
    MalS = case_when(
      Leptosphaeriaceae >= 1 & Plenodomus >= 1 ~ "high",
      Leptosphaeriaceae >= 1 & Plenodomus == 0 ~ "low",
      TRUE ~ "uninfected"
    )) %>%
  select(Sample, DED, Phyto, MalS)

###now calculate the number of each infection
sum(classified_df$DED=="high")
num_positive= classified_df %>% 
  summarise(DED_high   = sum(DED == "high"),
            Phyto_high = sum(Phyto == "high"),
            MalS_high  = sum(MalS == "high"), 
            DED_low    = sum(DED == "low"),
            Phyto_low  = sum(Phyto == "low"),
            MalS_low   = sum(MalS == "low"))

###add the coords and year info
for_GIS=left_join(classified_df, coords, by = "Sample")
for_GIS=left_join(for_GIS, year, by = "Sample")
for_GIS=for_GIS %>% 
  filter(!is.na(Lat))
for_GIS
str(for_GIS)

###add the ploidy info
ploidy=read.csv("C:/Users/mquig/Desktop/DED/Data_analysis/disease/ELM_spreadsheet - ploidy_results.csv", header = T)
##rename  the column
ploidy=ploidy %>% rename("Sample" = Sample_ID)
##combine with the big dataset
for_GIS=left_join(for_GIS, ploidy, by = "Sample")

###remove individuals with unknown ploidy and missing coords
for_GIS=for_GIS %>% 
  na.omit(ploidy) %>% 
  filter(ploidy != "other")

###make one column with the coinfections
all_infections <- for_GIS %>%
  mutate(
    infection_status = case_when(
      DED %in% c("high", "low") & Phyto %in% c("high", "low") & MalS %in% c("high", "low") ~ "DED/Phytoplasma/MalSecco",
      DED %in% c("high", "low") & Phyto %in% c("high", "low") ~ "DED/Phytoplasma",
      DED %in% c("high", "low") & MalS %in% c("high", "low") ~ "DED/MalSecco",
      Phyto %in% c("high", "low") & MalS %in% c("high", "low") ~ "Phytoplasma/MalSecco",
      DED %in% c("high", "low") ~ "DED",
      Phyto %in% c("high", "low") ~ "Phytoplasma",
      MalS %in% c("high", "low") ~ "MalSecco",
      TRUE ~ "uninfected"))

#write.csv(all_infections, "C:/Users/mquig/Desktop/DED/Data_analysis/disease/updated_NA_locations/all_infections_status_final_NAupdate.csv")
```

## Number of Infections
```{r}
###DED
DED_infected=all_infections %>% 
  filter(DED != "uninfected")
nrow(DED_infected)
#56
##make csv for mapping
DED_infected_map=all_infections %>% 
  filter(DED != "uninfected")
#write.csv(DED_infected_map, "C:/Users/mquig/Desktop/DED/Data_analysis/disease/updated_NA_locations/DED_final.csv")

###Phytoplasma
phyto_infected=all_infections %>% 
  filter(Phyto != "uninfected")
nrow(phyto_infected)
#22
##make csv for mapping
phyto_infected_map=all_infections %>% 
  filter(Phyto != "uninfected")
#write.csv(phyto_infected_map, "C:/Users/mquig/Desktop/DED/Data_analysis/disease/updated_NA_locations/phyto_final.csv")

###Mal Secco
malsecco_infected=all_infections %>% 
  filter(MalS != "uninfected")
nrow(malsecco_infected)
#123
##make csv for mapping
mals_infected_map=all_infections %>% 
  filter(MalS != "uninfected")
#write.csv(mals_infected_map, "C:/Users/mquig/Desktop/DED/Data_analysis/disease/updated_NA_locations/malsecco_final.csv")
```
##Total number of infections/coinfections/tri-infections
```{r}
sum(all_infections$infection_status=="uninfected")
#385
sum(all_infections$infection_status=="DED")
#13
sum(all_infections$infection_status=="DED/Phytoplasma")
#4
sum(all_infections$infection_status=="Phytoplasma")
#7
sum(all_infections$infection_status=="Phytoplasma/MalSecco")
#4
sum(all_infections$infection_status=="MalSecco")
#80
sum(all_infections$infection_status=="DED/MalSecco")
#32
sum(all_infections$infection_status=="DED/Phytoplasma/MalSecco")
#7
```

### High vs. low confidence infections
```{r}
###DED
##high
sum(DED_infected$DED=="high")
#3
sum(DED_infected$DED=="low")
#53

##phyto
sum(phyto_infected$Phyto=="high")
#21
sum(phyto_infected$Phyto=="low")
#1

##malsecco
sum(malsecco_infected$MalS=="high")
#118
sum(malsecco_infected$MalS=="low")
#5
```

## Oldest Infections
I want to know the earliest record of infections based on this dataset.
```{r}
###identify oldest infection
##DED
DED_pos=for_GIS %>% 
  filter(DED!="uninfected")
min(DED_pos$Year)
#1903

##phyto
phyto_pos=for_GIS %>% 
  filter(Phyto!="uninfected")
min(phyto_pos$Year)
#1909

##MalS
MalS_pos=for_GIS %>% 
  filter(MalS!="uninfected")
min(MalS_pos$Year)
#1897
```

### Cumulative number of infections over time
```{r}
### cumulative dataframe
cumulative_infect <- all_infections %>%
  mutate(
    DED   = ifelse(DED %in% c("high","low"), 1, 0),
    ElmYellows = ifelse(Phyto %in% c("high","low"), 1, 0),
    MalSecco  = ifelse(MalS %in% c("high","low"), 1, 0),
    Total = 1   # every sample counts toward cumulative total
  )

### convert to long format
cumulative_long <- cumulative_infect %>%
  select(Year, DED, ElmYellows, MalSecco, Total) %>%
  pivot_longer(cols = -Year, names_to = "Disease", values_to = "Infected")

### calculate the infections
cumulative_df <- cumulative_long %>%
  group_by(Disease, Year) %>%
  summarise(count = sum(Infected), .groups = "drop") %>%
  group_by(Disease) %>%
  arrange(Year) %>%
  mutate(cumulative = cumsum(count))

### graph it
ggplot(cumulative_df, aes(x = Year, y = cumulative, color = Disease)) +
  geom_line(size = 1) +
  #geom_point(size = 2) +
  scale_color_manual(values = c("purple4", "cornflowerblue", "springgreen4", "grey70")) +
  xlim(1890, 2025) +
  ylim(0, 550) +
  labs(
    #title = "Cumulative Infections Over Time",
    x = "Year",
    y = "Cumulative Number of Infected Individuals") +
  theme_classic() +
  theme(
    legend.position = c(0.035, 0.95),   # relative coordinates (x, y) inside plot
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = "black") # box around legend
  )
```
###Cumulative percentage of infected individuals over time
So today I presented at the Hipp lab meeting, and there were some concerns with this graph. Which is understandable, I thought the platau was weird too. Now, I'm going to try it again and see if anything changes.
```{r}
all_infections$Year=as.numeric(all_infections$Year)
cumsum_diseases=all_infections %>% 
   mutate(
    DED   = ifelse(DED   %in% c("high","low"), 1, 0),
    ElmYellows = ifelse(Phyto %in% c("high","low"), 1, 0),
    MalSecco  = ifelse(MalS  %in% c("high","low"), 1, 0)) %>% 
  arrange(Year) %>% 
  mutate(num_samples = 1) 

cumsum_diseases2=cumsum_diseases %>% 
  arrange(Year) %>% 
  group_by(Year) %>%
  summarise(
    sum_DED = sum(DED),
    sum_EY  = sum(ElmYellows),
    sum_MalS = sum(MalSecco),
    num_samples = sum(num_samples),
    .groups = "drop")

cumsum_diseases3=cumsum_diseases2 %>% 
  mutate(cumsum_DED=cumsum(sum_DED)) %>% 
  mutate(cumsum_EY=cumsum(sum_EY)) %>% 
  mutate(cumsum_MalS=cumsum(sum_MalS)) %>% 
  mutate(num_samples = cumsum(num_samples)) %>% 
  select(Year, cumsum_DED, cumsum_EY, cumsum_MalS, num_samples)

percent_infect=cumsum_diseases3 %>% 
  mutate(DED=(cumsum_DED/num_samples)*100) %>% 
  mutate(ElmYellows=(cumsum_EY/num_samples)*100) %>% 
  mutate(MalSecco=(cumsum_MalS/num_samples)*100) %>% 
  select(Year, DED, ElmYellows, MalSecco)

percent_graph=percent_infect %>% 
  pivot_longer(
    cols = c(DED, ElmYellows, MalSecco),
    names_to = "Disease",
    values_to = "percent_infected"
  )

ggplot(percent_graph, aes(x = Year, y = percent_infected, color = Disease)) +
    geom_vline(xintercept = c(1930, 1880, 2024), color = c("purple4", "springgreen4", "cornflowerblue"), linetype = "dashed", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  #geom_point(size = 2) +
  labs(
    #title = "Cumulative percentage of infected individuals over time",
    x = "Year of Sample Collection",
    y = "Cumulative percentage of Infected Individuals") +
  theme_classic() +
  #ylim(0, 31) +
  theme(
    legend.position = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = "black")) +
  scale_color_manual(values = c("purple4", "springgreen4", "cornflowerblue"))
```
After showing this graph to some people, they don't like it. Everyone thinks we need to somehow show the number of samples. 

###Stacked bar plot for each disease
The main recommendation was a stacked barplot over time. The years would have to be binned, so that is challenge #1. The other challenge is the coinfections. Jeremie recommended that I do one graph for each disease, then put it in the figure along the maps.

I don't think I'm going to like the outcome, but let's try it.

####DED
```{r}
###make the binned dataframe
binned_20yr <- cumsum_diseases2 %>%
  mutate(
    Year_bin = case_when(
      Year >= 1842 & Year <= 1860 ~ "1842-1860",
      Year >= 1861 & Year <= 1880 ~ "1861-1880",
      Year >= 1881 & Year <= 1900 ~ "1881-1900",
      Year >= 1901 & Year <= 1920 ~ "1901-1920",
      Year >= 1921 & Year <= 1940 ~ "1921-1940",
      Year >= 1941 & Year <= 1960 ~ "1941-1960",
      Year >= 1961 & Year <= 1980 ~ "1961-1980",
      Year >= 1981 & Year <= 2000 ~ "1981-2000",
      Year >= 2000 & Year <= 2020 ~ "2000-2020",
      Year >= 2020 & Year <= 2024 ~ "2020-2024")) %>%
  group_by(Year_bin) %>%
  summarise(
    sum_DED   = sum(sum_DED),
    sum_EY    = sum(sum_EY),
    sum_MalS  = sum(sum_MalS),
    num_samples = sum(num_samples),
    .groups = "drop"
  )

###DED
##make dataframe
DED_binned=binned_20yr %>% 
  mutate(num_uninfected=num_samples - sum_DED) %>% 
  select(Year_bin, sum_DED, num_uninfected) %>% 
  rename(infected=sum_DED) %>% 
  rename(uninfected=num_uninfected) %>% 
  rename(Year=Year_bin)
##pivot longer
DED_binned_long <- DED_binned %>%
  pivot_longer(
    cols = c(uninfected, infected),
    names_to = "Category",
    values_to = "Count")
##plot it
ggplot(DED_binned_long, aes(x = Year, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(
    values = c("uninfected" = "grey70", "infected" = "purple4")) +
  labs(
    x = "Year of Sample Collection",
    y = "Number of Samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.05, 0.95),
        legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", color = "black"))

###elm yellows
##make dataframe
EY_binned=binned_20yr %>% 
  mutate(num_uninfected=num_samples - sum_EY) %>% 
  select(Year_bin, sum_EY, num_uninfected) %>% 
  rename(infected=sum_EY) %>% 
  rename(uninfected=num_uninfected) %>% 
  rename(Year=Year_bin)
##pivot longer
EY_binned_long <- EY_binned %>%
  pivot_longer(
    cols = c(uninfected, infected),
    names_to = "Category",
    values_to = "Count")
##plot it
ggplot(EY_binned_long, aes(x = Year, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(
    values = c("uninfected" = "grey70", "infected" = "springgreen4")) +
  labs(
    x = "Year of Sample Collection",
    y = "Number of Samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.05, 0.95),
        legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", color = "black"))

###mal secco
##make dataframe
MalS_binned=binned_20yr %>% 
  mutate(num_uninfected=num_samples - sum_MalS) %>% 
  select(Year_bin, sum_MalS, num_uninfected) %>% 
  rename(infected=sum_MalS) %>% 
  rename(uninfected=num_uninfected) %>% 
  rename(Year=Year_bin)
##pivot longer
MalS_binned_long <- MalS_binned %>%
  pivot_longer(
    cols = c(uninfected, infected),
    names_to = "Category",
    values_to = "Count")
##plot it
ggplot(MalS_binned_long, aes(x = Year, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(
    values = c("uninfected" = "grey70", "infected" = "cornflowerblue")) +
  labs(
    x = "Year of Sample Collection",
    y = "Number of Samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.05, 0.95),
        legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", color = "black"))
```
I was right... I hate this :)

## Coinfections
Apparently, very little is known about coinfections. I want to check and see if any of these trees are infected with 2+ diseases at once. *this is not correct*
```{r}
##DED and phytoplasma
DED_phyto=for_GIS %>% 
  filter(DED!="uninfected") %>% 
  filter(Phyto!="uninfected")
nrow(DED_phyto)
#11 co-infections

##DED and MalS
DED_MalS=for_GIS %>% 
  filter(DED!="uninfected") %>% 
  filter(MalS!="uninfected")
nrow(DED_MalS)
#40 co-infection

##Phyto and MalS
Phyto_MalS=for_GIS %>% 
  filter(MalS!="uninfected") %>% 
  filter(Phyto!="uninfected")
nrow(Phyto_MalS)
#11

##all 3
all_three=for_GIS %>% 
  filter(DED!="uninfected") %>% 
  filter(Phyto!="uninfected") %>% 
  filter(MalS!="unifected")
nrow(all_three)
#11
```
## Incorportate ploidy
I also want to look at how disease is present across each ploidy.
```{r}
###make dataframes for each cytotype
diploid_infections= all_infections %>% 
  filter(ploidy == "2n")

polyploid_infections= all_infections %>% 
  filter(ploidy == "4n")
```
### Number of infections
```{r}
####diploids
###DED
diploid_DED_infected=diploid_infections %>% 
  filter(DED != "uninfected")
nrow(diploid_DED_infected)
#9
sum(diploid_DED_infected$DED=="high")
sum(diploid_DED_infected$DED=="low")

###phyto
diploid_phyto_infected=diploid_infections %>% 
  filter(Phyto != "uninfected")
nrow(diploid_phyto_infected)
#2
sum(diploid_phyto_infected$Phyto=="high")
sum(diploid_phyto_infected$Phyto=="low")

###mal secco
diploid_mals_infected=diploid_infections %>% 
  filter(MalS != "uninfected")
nrow(diploid_mals_infected)
#18
sum(diploid_mals_infected$MalS=="high")
sum(diploid_mals_infected$MalS=="low")

```

```{r}
####polyploid
###DED
polyploid_DED_infected=polyploid_infections %>% 
  filter(DED != "uninfected")
nrow(polyploid_DED_infected)
#47
sum(polyploid_DED_infected$DED=="high")
sum(polyploid_DED_infected$DED=="low")

###phyto
polyploid_phyto_infected=polyploid_infections %>% 
  filter(Phyto != "uninfected")
nrow(polyploid_phyto_infected)
#20
sum(polyploid_phyto_infected$Phyto=="high")
sum(polyploid_phyto_infected$Phyto=="low")

###mal secco
###DED
polyploid_mals_infected=polyploid_infections %>% 
  filter(MalS != "uninfected")
nrow(polyploid_mals_infected)
#105
sum(polyploid_mals_infected$MalS=="high")
sum(polyploid_mals_infected$MalS=="low")
```
#### random filtering
```{r, include=FALSE}
NA_samples=all_infections %>% 
  filter(str_detect(Sample, "NA"))

extractions=read.csv("Elm DNA Extraction Log - Field Collections.csv", header = TRUE)
extractions=extractions %>% 
  rename(Sample=CBG.Sample.ID)

NA_samples2=left_join(NA_samples, extractions, by = "Sample")

NA_for_export=NA_samples2 %>% 
  select(Sample, Outside.ID, DED, Phyto, MalS, Lat, Long, ploidy, infection_status, Origin)

write.csv(NA_for_export, "C:/Users/mquig/Downloads/NA_samples.csv")
```

## Upset plot
Suzy recommended I try an upset plot to visualize the disease presence over time. I think this would be a great way to show the infection rates over the years and the number of samples collected over time. 

I am using this tutorial (https://albert-rapp.de/posts/ggplot2-tips/26_upset_charts/26_upset_charts.html), specifically the part where we make each plot independently then combine them.

### Plot 1) number of samples over time
For this we have two options, the cumulative number of samples over time, or just the number of samples collected during that time frame.
```{r, include=TRUE}
###make a dataframe with 10 year intervals
binned_10yr <- cumsum_diseases2 %>%
  mutate(
    Year_bin = case_when(
      Year >= 1842 & Year <= 1850 ~ "1842-1850",
      Year >= 1851 & Year <= 1860 ~ "1851-1860",
      Year >= 1861 & Year <= 1870 ~ "1861-1870",
      Year >= 1871 & Year <= 1880 ~ "1871-1880",
      Year >= 1881 & Year <= 1890 ~ "1881-1890",
      Year >= 1891 & Year <= 1900 ~ "1891-1900",
      Year >= 1901 & Year <= 1910 ~ "1901-1910",
      Year >= 1911 & Year <= 1920 ~ "1911-1920",
      Year >= 1921 & Year <= 1930 ~ "1921-1930",
      Year >= 1931 & Year <= 1940 ~ "1931-1940",
      Year >= 1941 & Year <= 1950 ~ "1941-1950",
      Year >= 1951 & Year <= 1960 ~ "1951-1960",
      Year >= 1961 & Year <= 1970 ~ "1961-1970",
      Year >= 1971 & Year <= 1980 ~ "1971-1980",
      Year >= 1981 & Year <= 1990 ~ "1981-1990",
      Year >= 1991 & Year <= 2000 ~ "1991-2000",
      Year >= 2001 & Year <= 2010 ~ "2001-2010",
      Year >= 2011 & Year <= 2020 ~ "2011-2020",
      Year >= 2021 & Year <= 2025 ~ "2021-2025")) %>%
  group_by(Year_bin) %>%
  summarise(
    sum_DED   = sum(sum_DED),
    sum_EY    = sum(sum_EY),
    sum_MalS  = sum(sum_MalS),
    num_samples = sum(num_samples),
    .groups = "drop") %>% 
  mutate(cumsum=cumsum(num_samples))

### plot number of samples
ggplot(binned_10yr, aes(x = Year_bin, y = num_samples)) +
  geom_col() +
    labs(
    x = "Year of Sample Collection",
    y = "Number of Samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.05, 0.95),
        legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", color = "black"))

### how about adding a line for cumulative number of samples?
ggplot(binned_10yr, aes(x = Year_bin)) + 
  geom_col(aes(y = num_samples), fill = "grey70") + 
  geom_line(aes(y = cumsum), group = 1, color = "cornflowerblue", linewidth = 1) + 
  geom_point(aes(y = cumsum), color = "cornflowerblue", size = 2) + 
  #scale_y_continuous( name = "Number of Samples", sec.axis = sec_axis(~ ., name = "Cumulative Samples") ) + 
  labs(x = "Year of Sample Collection", y = "Number of Samples") + 
  theme_classic() + 
  theme( axis.text.x = element_text(angle = 45, hjust = 1) )


#plot cumulative number of samples
ggplot(binned_10yr, aes(x = Year_bin, y = cumsum)) +
  geom_col() +
    labs(
    x = "Year of Sample Collection",
    y = "Cumulative Number of Samples") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.05, 0.95),
        legend.justification = c("left", "top"),
        legend.background = element_rect(fill = "white", color = "black"))

### add a symbol denoting the first records of infection
first_id=tibble(
  disease=c("DED", "ElmYellows", "MalSecco"),
  Year_bin=c("1931-1940", "1881-1890", "2021-2025"))
first_id2=first_id %>% 
  left_join(binned_10yr, by = "Year_bin")

## graph it
ggplot(binned_10yr, aes(x = Year_bin, y = num_samples)) +
  geom_col(fill="grey30") +
  geom_point(data = first_id2,
             aes(x = Year_bin, y = num_samples + 7, color = disease),
             size = 5, shape = 17) +
  labs(x = "Year of Sample Collection", y = "Number of Samples", 
       color = "Earliest Record") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = "black"))+
  scale_color_manual(values = c("purple4", "springgreen4", "cornflowerblue"))

```
### Plot 2) Infection status
Now I want to do the little circles for infection status.
```{r}
### convert to long format
binned_10yr_long <- binned_10yr %>%
  rename(DED=sum_DED) %>% 
  rename(ElmYellows=sum_EY) %>% 
  rename(MalSecco=sum_MalS) %>% 
  pivot_longer(cols = c(DED, ElmYellows, MalSecco),
               names_to = "disease",
               values_to = "count") %>%
  mutate(filled = count >= 1)

ggplot(binned_10yr_long, aes(x = Year_bin, y = disease)) +
  geom_point(aes(fill = filled),
             shape = 21, size = 4) +
  scale_fill_manual(values = c("white", "black")) +
  theme_void() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "", y = "")

### graph it
binned_10yr_long <- binned_10yr_long %>%
  mutate(fill_color = ifelse(filled,
                             disease,
                             "none"))

ggplot(binned_10yr_long, aes(x = Year_bin, y = disease)) +
  geom_point(
    aes(fill = fill_color),
    shape = 21, size = 6, color = "black"
  ) +
  scale_fill_manual(values = c(
    "DED"  = "purple4",
    "ElmYellows"   = "springgreen4",
    "MalSecco" = "cornflowerblue",
    "none" = "white"
  )) +
  theme_void() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(x = "", y = "")
```
### Plot 3) Sideways bar graph of infected individuals
```{r}
#make the dataframe
total_infect=binned_10yr %>% 
  summarise(DED = sum(sum_DED), 
            ElmYellows = sum(sum_EY, na.rm = TRUE), 
            MalSecco = sum(sum_MalS, na.rm = TRUE)) %>%
  pivot_longer(everything(), names_to = "disease", values_to = "total_infections")

#graph it
ggplot(total_infect, aes(x = total_infections, y = disease)) +
  geom_col(fill = c("purple4", "springgreen4", "cornflowerblue")) +
  theme_classic() +
  labs(x = "Number of Infections", y="") 
```
