---
title: "Mal Secco ITS Phylogeny"
author: "m quigg"
date: "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Phylogeny of Mal Secco
### The goal here is to make a phylogeny of the gene regions originating from the Mal Secco pathogens. I used all of the "infected" individuals and generated the phylogeny from that.

## Prep
```{r include=FALSE}
### grab your library card ###
library(ggtree)
library(ape)
library(treeio)
library(tidyverse)
library(janitor)
library(Biostrings)
library(patchwork)
library(tidytree)


### set working directory ###
setwd("C:/Users/mquig/Desktop/DED/Data_analysis/disease_phylo/mal_secco")
```

## Import the Data
```{r}
###the important stuff
tree_ITS = treeio::read.tree("aligned.fa.treefile")

###the fun stuff
coords=read.csv("/Users/mquig/Desktop/DED/Data_analysis/disease/master_list_NA_updated.csv", header = T)
year=read.csv("/Users/mquig/Desktop/DED/Data_analysis/disease/samples_and_year.csv", header = T)

###clean up the fun stuff
year = year %>% 
  dplyr::rename("SampleID" = ind)
coords = coords %>% 
  select(Sample.ID, Lat.NA_corrected., Long.NA_corrected.) %>% 
  dplyr::rename("SampleID" = Sample.ID, "Latitude" = Lat.NA_corrected., "Longitude" = Long.NA_corrected.)
```

## ITS Gene Region: Create the Phylogeny
### Add a Root
```{r}
rooted_tree_ITS=root(tree_ITS, outgroup = "PQ351261")
```

### Graph it
#### Basic tree
```{r}
ggtree(tree_ITS, branch.length = "none", layout = "circular")
```
### Add Metadata
```{r, include=FALSE}
# Fix the treetips creation - remove the duplicate label column
treetips_ITS = as.data.frame(tree_ITS$tip.label)
names(treetips_ITS)[1] = "label"  # keep it as 'label' directly

treetips_ITS = treetips_ITS %>% 
  mutate(sample_id = label) %>%   # keep original for joining
  separate(col="sample_id", into=c("SampleID", "batch"), sep="_") %>%
  select(-batch)

# Join metadata
treetips_meta_ITS = treetips_ITS %>%
  left_join(coords, by="SampleID") %>%
  left_join(year, by="SampleID")

# Check it looks right
head(treetips_meta_ITS)
colnames(treetips_meta_ITS)
```

### Make fun phylogenies!
```{r}
###add tip labels
ggtree(tree_ITS, branch.length="none") +
  geom_tiplab(color='navy', size=2) +
  geom_nodelab(size=2, color="red")
```

## Tree with both ITS and TUB2
Now I want to make a tree with both the ITS genes and the TUB2 genes.
```{r}
tree_both = treeio::read.tree("concatenated_MalS.fasta.treefile")

###add a root
rooted_tree_both=root(tree_both, outgroup = "PQ351261_OQ412812")
```

### Graph it
#### Basic tree
```{r}
ggtree(rooted_tree_both, branch.length = "none", layout = "circular")

ggtree(rooted_tree_both, branch.length="none") +
  geom_tiplab(color='navy', size=2)

###make a tree with bootstrap numbers
ggtree(rooted_tree_both, branch.length='none', linewidth=0.5) +
      geom_nodelab(size=2, color="red") +
      geom_tiplab(color='navy', size=2)
```
OH MY GOD!!! It looks like we have two strains??? There are def two clades emerging, and that initial split has a high bootstrap level. Within clades is a little messy, with low bootstrpping but that initial split seems pretty confident.

Now I want to compare spatially, and temporally.

### Add the metadata
```{r, include=FALSE}
# Fix the treetips creation - remove the duplicate label column
treetips_both = as.data.frame(tree_both$tip.label)
names(treetips_both)[1] = "label"  # keep it as 'label' directly

treetips_both = treetips_both %>% 
  mutate(sample_id = label) %>%   # keep original for joining
  separate(col="sample_id", into=c("SampleID", "batch"), sep="_") %>%
  select(-batch)

# Join metadata
treetips_meta_both = treetips_both %>%
  left_join(coords, by="SampleID") %>%
  left_join(year, by="SampleID")

# Check it looks right
head(treetips_meta_both)
colnames(treetips_meta_both)
```


### Spatial
*a quick note: the coords here have the NatArb trees at the plot in Maryland. There are two possibilities, one is that they got infected at the NatArb, the other is that they were already infected before reaching the NatArb. It looks like both may be happening... There are two groups to NatArb trees, one in each clade!

#### Latitude
The goal is to make a phylogeny where the colors of the tips correspond to the latitude the sample was taken from.
```{r}
###create the color palette 
lat_min <- min(treetips_meta_both$Latitude, na.rm=TRUE)
lat_max <- max(treetips_meta_both$Latitude, na.rm=TRUE)

###graph the cladogram
tree_plot <- ggtree(rooted_tree_both, branch.length='none') %<+% treetips_meta_both +
    geom_tippoint(aes(color=Latitude), size=2) +
    scale_color_gradient(low="lightblue", high="darkblue", na.value="grey50")
tree_plot

###make the map
us_map <- map_data("state")

map_data_pts <- treetips_meta_both %>%
    filter(!is.na(Latitude), !is.na(Longitude))

map_plot <- ggplot() +
    geom_polygon(data=us_map, aes(x=long, y=lat, group=group),
                 fill="lightgrey", color="white", linewidth=0.3) +
    geom_point(data=map_data_pts,
               aes(x=Longitude, y=Latitude, color=Latitude),
               size=3, alpha=0.8) +
    scale_color_gradient(
        low="lightblue", high="darkblue",
        name="Latitude"
    ) +
    coord_fixed(1.3, xlim=c(-125, -65), ylim=c(24, 50)) +
    theme_void() +
    theme(panel.border=element_rect(color="black", fill=NA, linewidth=0.5), legend.position="none")
map_plot

###combine
tree_plot + inset_element(map_plot, left=0.0, bottom=0.7, right=0.3, top=1.0)
```

####Longitude
The goal is to make a phylogeny where the colors of the tips correspond to the longitude the sample was taken from.
```{r}
###create the color palette 
long_min <- min(treetips_meta_both$Longitude, na.rm=TRUE)
long_max <- max(treetips_meta_both$Longitude, na.rm=TRUE)

###graph the cladogram
tree_plot_long <- ggtree(rooted_tree_both, branch.length='none') %<+% treetips_meta_both +
    geom_tippoint(aes(color=Longitude), size=2) +
    scale_color_gradient(low="lightgreen", high="darkgreen", na.value="grey50")
tree_plot_long

###make the map
us_map <- map_data("state")

map_data_pts <- treetips_meta_both %>%
    filter(!is.na(Latitude), !is.na(Longitude))

map_plot_long <- ggplot() +
    geom_polygon(data=us_map, aes(x=long, y=lat, group=group),
                 fill="lightgrey", color="white", linewidth=0.3) +
    geom_point(data=map_data_pts,
               aes(x=Longitude, y=Latitude, color=Longitude),
               size=3, alpha=0.8) +
    scale_color_gradient(
        low="lightgreen", high="darkgreen",
        name="Longitude"
    ) +
    coord_fixed(1.3, xlim=c(-125, -65), ylim=c(24, 50)) +
    theme_void() +
    theme(panel.border=element_rect(color="black", fill=NA, linewidth=0.5), legend.position="none")
map_plot_long

###combine
tree_plot_long + inset_element(map_plot_long, left=0.0, bottom=0.7, right=0.3, top=1.0)
```

####Year
The goal is to make a phylogeny where the colors of the tips correspond to the longitude the sample was taken from.
```{r}
###create the color palette 
year_min <- min(treetips_meta_both$year, na.rm=TRUE)
year_max <- max(treetips_meta_both$year, na.rm=TRUE)

###graph the cladogram
tree_plot_year <- ggtree(rooted_tree_both, branch.length='none') %<+% treetips_meta_both +
    geom_tippoint(aes(color=year), size=2) +
    scale_color_gradient(low="#FF99CC", high="#660033", na.value="grey50")
tree_plot_year

###make the map
us_map <- map_data("state")

map_data_pts <- treetips_meta_both %>%
    filter(!is.na(Latitude), !is.na(Longitude))

map_plot_year <- ggplot() +
    geom_polygon(data=us_map, aes(x=long, y=lat, group=group),
                 fill="lightgrey", color="white", linewidth=0.3) +
    geom_point(data=map_data_pts,
               aes(x=Longitude, y=Latitude, color=year),
               size=3, alpha=0.8) +
    scale_color_gradient(
        low="#FF99CC", high="#660033",
        name="Year"
    ) +
    coord_fixed(1.3, xlim=c(-125, -65), ylim=c(24, 50)) +
    theme_void() +
    theme(panel.border=element_rect(color="black", fill=NA, linewidth=0.5), legend.position="none")
map_plot_year

###combine
tree_plot_year + inset_element(map_plot_year, left=0.0, bottom=0.7, right=0.3, top=1.0)
```
### Assign Clades
Ok so there are 2 clades. I need to assign each individual to one of the two main clades.
```{r}
###plot with clade numbers
ggtree(rooted_tree_both, branch.length='none', linewidth=0.5) +
    geom_text(aes(label=node), size=2, color="red") 
#clade 1 is 127 and clade 2 is 183

###extract individuals in each clade
clade1_tips <- extract.clade(rooted_tree_both, node=127)$tip.label
clade2_tips <- extract.clade(rooted_tree_both, node=183)$tip.label

###make a dataframe
clade1=data.frame(SampleID = clade1_tips)
clade1=clade1 %>% 
  mutate(clade = "1")

clade2=data.frame(SampleID = clade2_tips)
clade2=clade2 %>% 
  mutate(clade = "2")

clade_assignments=rbind(clade1, clade2)

###add coords
clade_coords=clade_assignments %>% 
  separate(SampleID, into = c("SampleID", "batch"), sep = "_") 
clade_coords=left_join(clade_coords, coords, by = "SampleID")
clade_coords_year=left_join(clade_coords, year, by = "SampleID")

###write csv to map in arcGIS
#write.csv(clade_coords_year, "MalS_clade_coords.csv")
```

### Final figure
I want to make a map where the shape of the points corresponds to the clade, and the color of the points is the year the sample was collected in. 
```{r}
###make the dataframe for the figure
treetips_final=clade_coords_year %>% 
  unite("SampleID", SampleID, batch, sep = "_")

###make the tree
tree_final=ggtree(rooted_tree_both, branch.length='none') %<+% treetips_final +
    geom_tippoint(aes(color=year, shape = clade), size=2) +
    scale_color_gradient(low="lightblue", high="darkblue", na.value="grey50")
tree_final

###make the map
us_map <- map_data("state")

map_data_pts <- treetips_final %>%
    filter(!is.na(Latitude), !is.na(Longitude))

map_plot_final <- ggplot() +
    geom_polygon(data=us_map, aes(x=long, y=lat, group=group),
                 fill="white", color="darkgrey", linewidth=0.3) +
    geom_point(data=map_data_pts,
               aes(x=Longitude, y=Latitude, color=year, shape = clade),
               size=3, alpha=1) +
    scale_color_gradient(
        low="lightblue", high="darkblue",
        name="Year"
    ) +
    coord_fixed(1.3, xlim=c(-125, -65), ylim=c(24, 50)) +
    theme_void() +
    theme(panel.border=element_rect(color="black", fill=NA, linewidth=0.5), legend.position="none")
map_plot_final

tree_final + inset_element(map_plot_final, left=0.0, bottom=0.7, right=0.3, top=1.0)
```

#### Boxplot: clades and years
I have an idea! What about a boxplot with both clades by year? That can show the variation of time!
```{r}
### make a dataframe
clade_year=clade_coords_year %>% 
  select(SampleID, clade, year) %>% 
  group_by(clade)

###graph it
ggplot(clade_year, aes(clade, year)) +
  geom_boxplot(fill = c("darkblue", "lightblue")) +
  theme_classic() +
  ylab("Year") + xlab("Clade") +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 1))
```

```{r}
###make the dataframe for the figure
treetips_final=clade_coords_year %>% 
  unite("SampleID", SampleID, batch, sep = "_")

###make the tree
tree_clade_color=ggtree(rooted_tree_both, branch.length='none') %<+% treetips_final +
    geom_tippoint(aes(color=clade), size=2) +
    scale_color_manual(values = c("1" ="darkblue", "2" = "lightblue", na.value="grey50"))
tree_clade_color

###make the map
us_map <- map_data("state")

map_data_pts <- treetips_final %>%
    filter(!is.na(Latitude), !is.na(Longitude))

map_plot_clade_color <- ggplot() +
    geom_polygon(data=us_map, aes(x=long, y=lat, group=group),
                 fill="white", color="black", linewidth=0.3) +
    geom_point(data=map_data_pts,
               aes(x=Longitude, y=Latitude, color=clade),
               size=3, alpha=1) +
    scale_color_manual(values=c("1"="darkblue", "2"="lightblue")) +  # <-- color not fill
    coord_fixed(1.3, xlim=c(-125, -65), ylim=c(24, 50)) +
    theme_void() +
    theme(panel.border=element_rect(color="black", fill=NA, linewidth=0.5), 
          legend.position="none")

tree_clade_color + inset_element(map_plot_clade_color, left=0.0, bottom=0.7, right=0.3, top=1.0)
```
